---
title: "BCB420 Assignemnt 1"
author: "George Ly"
output:
  html_document:
    df_print: paged
---

I have chosen to work on a data-set exploring the differences in gene expression
of patients with asthma in response to the SARS-CoV-2 virus. 

I have chosen this data-set to work with as asthma is a problem myself and others 
in my family have struggled with in the past. With asthma being a respiratory 
disease, I am curious to know what possible effects it could have in conjunction 
with another respiratory illness such as COVID-19.

```{r, message=FALSE}
if (!file.exists("GSE158752_IMSA_BEC-RAW_Counts.txt")) {
  GEOquery::getGEOSuppFiles("GSE158752",makeDirectory = FALSE)
  gunzip("GSE158752_IMSA_BEC-RAW_Counts.txt.gz")
}
```

```{r}
rawCounts = read.table("GSE158752_IMSA_BEC-RAW_Counts.txt",header=TRUE, 
                       row.names = 1, check.names = FALSE)
nrow(rawCounts)
rownames(rawCounts)[1:10]
summarizedGeneCounts <- sort(table(rownames(rawCounts)),decreasing = TRUE)
summarizedGeneCounts[1:10]
```

There are 26234 genes represented within the expression data-set. 
The gene names (which were set to the row names) are already in HUGO format.
Thus there will be no mapping done here. Every gene represented in the data-set
is only represented once.

```{r}
ncol(rawCounts)
colnames(rawCounts)[1:10]
```
There are 67 samples within this experiment. 

The samples are split into three groups, the control group "Healthy" with no 
asthma, and two experimental groups with "Mild" Asthma and "Severe" Asthma. These 
are represented respectively by "H", "M", & "S".

```{r}
sum(unlist(lapply(colnames(rawCounts),grep,pattern="_H")))
sum(unlist(lapply(colnames(rawCounts),grep,pattern="_M")))
sum(unlist(lapply(colnames(rawCounts),grep,pattern="_S")))
samples <- data.frame(lapply(colnames(rawCounts),
                      FUN=function(x){unlist(strsplit(x, split = "\\_"))[c(1,3)]}))
colnames(samples) <- colnames(rawCounts)
rownames(samples) <- c("Patient","Asthma_severity")
samples <- data.frame(t(samples))
```
There are 17 healthy samples, and 25 samples with mild and severe asthma.

```{r}
cpms = edgeR::cpm(rawCounts)
keep = rowSums(cpms >1) >=67
countsFiltered = rawCounts[keep,]
nrow(countsFiltered)
```
After calculating for counts-per-million (CPM) using the edgeR package, I filtered
out genes with less than 1 CPM per sample. This leaves 12865 genes left in the data-set after 
13369 genes were filtered out.

```{r}
filteredDataMatrix <- as.matrix(countsFiltered)
d = edgeR::DGEList(counts=filteredDataMatrix, group=samples$Asthma_severity)
d = edgeR::calcNormFactors(d)
TMMCounts = edgeR::cpm(d)
```

I chose to use the TMM normalization method in favour of distribution based methods
as there are less control samples when compared to the two experimental sample sets.

```{r}
originalData <- log2(edgeR::cpm(countsFiltered))
boxplot(originalData, xlab = "Samples", ylab = "log2 (CPM)", las=2, xaxt='n', main = "Before Normalization")
abline(h = median(apply(originalData, 2, median)), col = "red")

normData <- log2(TMMCounts)
boxplot(normData, xlab = "Samples", ylab = "log2 (CPM)", las=2, xaxt='n', main = "After Normalization")
abline(h = median(apply(normData, 2, median)), col = "red")
```


As expected the normalization does not change the results of the box-plot very much.
```{r}
limma::plotMDS(d, labels=rownames(samples),
        col = c("darkgreen","blue","red")[factor(samples$Asthma_severity)], main = "MDS Plot")
```


The MDS shows samples with healthy and moderate cases of Asthma being clustered 
close together. Severe cases of asthma are more likely to be farther away from
the main cluster.